---
title: "Morning Practical"
output: html_notebook
---

# Ozone Modelling

- Get the data

```{r get-ozone}
ozone <- read.table("../workshop_JGI_2018/exercises/data/ozone.txt")
```

- Look at the data

```{r look-at-ozone}
pairs(ozone)
```

- Load required packages

```{r packages}
library(mgcViz)
```

- Fit a Gaussian GAM with 03 as a response, where log(E(03)) is given by the sum of smooth functions

```{r fit-ozone}
ozone_gam <- gamV(O3 ~ s(vh) + s(wind) + s(humidity) + s(temp) + s(ibh) + s(dpg) + s(ibt) + s(vis) + s(doy), 
             family = gaussian(link = log),
             data = ozone,
             aViz = list(nsim = 50))

print(plot(ozone_gam, allTerms = TRUE), pages = 1) # plot() calls plot.gamViz(). See ?plot.gamViz
```

- Check the model residuals

```{r check-model-residuals}
qq(ozone_gam)
check(ozone_gam)
```

- Refit using a gamma distribution, does this improve fit?

```{r gamma-ozone-fit}
ozone_gam_gamma <- gamV(O3 ~ s(vh) + s(wind) + s(humidity) + s(temp) + s(ibh) + s(dpg) + s(ibt) + s(vis) + s(doy), 
             family = Gamma(link = log),
             data = ozone,
             aViz = list(nsim = 50))

check(ozone_gam_gamma)
```

- Fit alternative model with an indentity-link

```{r gamma-identity-link}
ozone_gam_gamma_id <- gamV(O3 ~ s(vh) + s(wind) + s(humidity) + s(temp) + s(ibh) + s(dpg) + s(ibt) + s(vis) + s(doy), 
             family = Gamma(link = identity),
             data = ozone,
             aViz = list(nsim = 50))

AIC(ozone_gam_gamma, ozone_gam_gamma_id)
```

- Plot the smoothed effects

```{r plot-smoothed-effects}
print(plot(ozone_gam_gamma, allTerms = TRUE), pages = 1) 

summary(ozone_gam_gamma)
```

- Simplify the model using p value selection (bad)

```{r p-value restricted model}
ozone_gam_gamma_rest <- gamV(O3 ~ s(vh) + s(wind) + s(temp) + s(ibh) + s(dpg) + s(vis) + s(doy), 
             family = Gamma(link = identity),
             data = ozone,
             aViz = list(nsim = 50))

AIC(ozone_gam_gamma_id, ozone_gam_gamma_rest)
```


- Interpret the smoothed effects, Q example of interpreting this?

```{r interpret-effects}
print(plot(ozone_gam_gamma_rest, allTerms = TRUE), pages = 1) 
```

# Bone mineral density modelling

- Get the calcium data

```{r get-cal-data}
load("../workshop_JGI_2018/exercises/data/calcium.rda")
```

- Fit a simple model with `bmd` as the reponse and linear effects for `age` and `groups`.

```{r fit-baseline-cal-model}
linear_gam <- gamV(bmd ~ group + age, data = calcium, aViz = list(nsim = 100))

summary(linear_gam)
```

- Check residuals if to see in mean depends on the subject

```{r check-baseline-residuals}
check1D(linear_gam, calcium$person) + l_gridCheck1D(mean)
```

- Add random effect for subjects

```{r add-random-effect}
cal_rf_per <- gamV(bmd ~ group + age + s(person, bs = "re"), 
                   data = calcium, 
                   aViz = list(nsim = 100))
check1D(cal_rf_per, calcium$person) + l_gridCheck1D(mean)
summary(cal_rf_per)
```

- Add a smooth effect for age and compare using AIC

```{r smooth-age}
col_age_smooth <- gamV(bmd ~ group + s(age) + s(person, bs = "re"), data = calcium)

print(plot(col_age_smooth, allTerms = T), pages = 1)
```

- Check to see if age impact is different between age group using a by factor.

```{r within-group-age}
col_age_smooth_by_age <- gamV(bmd ~ group + s(age, by = group) + s(person, bs = "re"), data = calcium)

print(plot(col_age_smooth_by_age, allTerms = T), pages = 1)
```

- Look at model differences

```{r plotdiffs}
plotDiff(sm(col_age_smooth_by_age, 1),
         sm(col_age_smooth_by_age, 2)) + l_fitLine() + l_ciLine()
```

# Forecasting electricity demand on GEFCOM2014 data

- Get the GEF com 2014 data

```{r get-gef-data}
load("../workshop_JGI_2018/exercises/data/gefcom_small.rda")
```

- Fit a Guassian GAM, with smooth effects for some variables and parameteric effects for others.


```{r fit-baseline-gef-model}
baseline_gef <- gamV(NetDemand ~ NetDemand.24 + Dow + Trend + s(wM) + s(wM_s95) + s(Posan, bs = "cc"), 
             data = gefcom_small, aViz = list(nsim = 50))

print(plot(baseline_gef, allTerms = TRUE), pages = 1)
```

- Check the mean of residuals depending on Trend. 

```{r check-trend-residuals}
check1D(baseline_gef, "Trend") + l_gridCheck1D()
```

```{r check-k-values}
check(baseline_gef)
```

- Refit the model the model with a smoothed effect for trend and a increased basis dimension for all spline variables. Compare via AIC and check residuals. Recheck basis dimensions using check.

```{r}
gef_smooth <- gamV(NetDemand ~ NetDemand.24 + Dow + s(Trend, k = 6) + s(wM, k = 20) + 
             s(wM_s95, k = 15) + s(Posan, k = 15, bs = "cc"),
             data = gefcom_small, 
             aViz = list(nsim = 50))

AIC(baseline_gef, gef_smooth)

check1D(gef_smooth, "Trend") + l_gridCheck1D()

check(gef_smooth)
```

- Check qq plot

```{r gef-qq}
qq(gef_smooth)
```

- Refit using a scaled students T distribution.

```{r scaled-student}
gef_scat <- gamV(NetDemand ~ NetDemand.24 + Dow + s(Trend, k = 6) + s(wM, k = 20) + 
             s(wM_s95, k = 15) + s(Posan, k = 15, bs = "cc"), data = gefcom_small, 
             family = scat,
             aViz = list(nsim = 50))

AIC(gef_smooth, gef_scat)

qq(gef_scat)
```

- Consider log link

```{r}
gef_scat_log <- gamV(NetDemand ~ NetDemand.24 + Dow + s(Trend, k = 6) + s(wM, k = 20) + 
             s(wM_s95, k = 15) + s(Posan, k = 15, bs = "cc"), 
             data = gefcom_small, 
             family = scat(link = log),
             aViz = list(nsim = 50))

AIC(gef_scat, gef_scat_log)

print(plot(gef_scat_log, allTerms = TRUE), pages = 1)
```