---
title: "Morning Practicals: Session 2"
output: html_notebook
---

# $CO_2$ Modelling

- Load the packages and data

```{r}
library(mgcv); library(gamair); 
data(co2s)
```

- Plot the data

```{r}
library(tidyverse)
ggplot(co2s, aes(c.month, co2)) +
  geom_line() +
  theme_minimal()
```

- Fit the GAM model

```{r}
baseline_gam <- gam(co2 ~ s(c.month, bs = "cr", k = 100), data = co2s)
```

- Make predictions

```{r}
pred_plot <- function(df, model) {
  
  preds <- predict(model ,data.frame(c.month=1:543,
                                month=c(rep(1:12,45),1:3)),
                   se=TRUE)
  
  out <- tibble(fit = preds$fit) %>% 
    mutate(
        ul = fit + 2*preds$se,
        ll = fit - 2*preds$se,
        c.month = 1:543
    )


  ggplot(df, aes(c.month, co2)) +
  geom_point(size = 1.2, alpha = 0.6) +
  geom_line(data = out, aes(y = fit), col = "red", alpha = 0.6) +
  geom_ribbon(data = out, 
              aes(x = c.month, y = fit, ymin = ll, ymax = ul),
              col = "red", alpha = 0.2) +
  theme_minimal() 

}

pred_plot(co2s, baseline_gam)
```

- Fit a model with smooth functions for c.month and month (i.e months with a season effect). Use a high diemnsion for f1 and a cyclic basis for f2. Need to set knots for month to force seasons to line up correctly.

```{r seasonal-gam}
seasonal_gam <- gam(co2 ~ s(c.month, bs = "cr",k = 50) + s(month,bs = "cc"),
                    data=co2s,
          knots=list(month=c(1,13)))


```

- Plot seasonal predictions

```{r seasonal-predictions}
pred_plot(co2s, seasonal_gam)
```

- Short term seasonality has been seperated from long term trends. This improves the extraplotion of longer term trends

# Individual electricity demand modelling solution

- Load the data and fit a basic model

```{r load-data-fit-gam}
load( "../workshop_JGI_2018/exercises/data/Ind_elect.rda") 

fit1 <- bamV(load ~ dow + ID + s(instant) + s(temp) + s(day), data = indElec, 
             aViz = list(nsim = 50))

print(plot(fit1, allTerms = TRUE), pages = 1)
```

- Get model summary

```{r model-summary}
summary(fit1)
```

- Check residuals across variables

```{r}
pl <- list()
pl[[1]] <- check2D(fit1, "instant", "ID") + l_gridCheck2D(bw = c(NA, 1))
pl[[2]] <- check2D(fit1, "temp", "ID") + l_gridCheck2D(bw = c(NA, 1))
pl[[3]] <- check2D(fit1, "day", "ID") + l_gridCheck2D(bw = c(NA, 5))

library(gridExtra)
grid.arrange(grobs = lapply(pl, "[[", "ggObj"), ncol = 2)
```

- Explore problematic customers. Looks like using storage heaters?

```{r}
in27 <- which(indElec$ID == 27)
par(mfrow = c(1, 2))
plot(indElec$instant[in27], indElec$load[in27])
plot(indElec$temp[in27], indElec$load[in27])
```

- Fit a model using BAM for individual variation

```{r fit-bam-model}
fit2 <- bamV(load ~ dow + ID + 
                    s(instant, by = ID, id = 1) + 
                    s(temp, by = ID, id = 2) + 
                    s(day, by = ID, id = 3), data = indElec, 
             aViz = list(nsim = 50), aGam = list(discrete = TRUE))

summary(fit2)
```

- Check AIC

```{r}
AIC(fit1, fit2)
```

- Repeat residual checks

```{r}
pl <- list()
pl[[1]] <- check2D(fit2, "instant", "ID") + l_gridCheck2D(bw = c(NA, 1))
pl[[2]] <- check2D(fit2, "temp", "ID") + l_gridCheck2D(bw = c(NA, 1))
pl[[3]] <- check2D(fit2, "day", "ID") + l_gridCheck2D(bw = c(NA, 5))

library(gridExtra)
grid.arrange(grobs = lapply(pl, "[[", "ggObj"), ncol = 2)
```

- Check basis functions. Basis for day and instant may be too small.

```{r}
check(fit2)
```

- Increase dimension and refit

```{r}
fit3 <- bamV(load ~ dow + ID + 
                    s(instant, by = ID, id = 1, k = 20) + 
                    s(temp, by = ID, id = 2) + 
                    s(day, by = ID, id = 3, k = 20), data = indElec, 
             aViz = list(nsim = 50), aGam = list(discrete = TRUE))

AIC(fit2, fit3)
```

- Check dimensions again

```{r}
check(fit3)
```

- Check relationship between individuals, load and relationship.

```{r}
check2D(fit3, indElec$load48SM, indElec$ID) + l_gridCheck2D(bw = c(NA, 0.1))
```

- Include interaction for smoothed lagged load and refit model

```{r}
fit4 <- bamV(load ~ dow + ID + 
                    s(instant, by = ID, id = 1, k = 20) + 
                    s(temp, by = ID, id = 2) + 
                    s(day, by = ID, id = 3, k = 20) + 
                    s(load48SM, by = ID, id = 4), data = indElec, 
             aViz = list(nsim = 50), aGam = list(discrete = TRUE))

AIC(fit3, fit4)
```

- Check predictions on individual customers

```{r}
par(mfrow = c(3, 3), mar = c(2.1, 4.1, 1.1, 2.1))
for(ID in 1:9){
  ids <- which( indElec$ID == ID )
  subD <- indElec[ids, ]
  subP <- fit4$fitted.values[ids]
  
  kk <- 1:200
  plot(subD$load[kk])
  lines(subP[kk], col = 2)
}

```

- Model is not going a good job at predicting peaks in demand in general. It is doing a particularly poor job for customer number 3.