---
title: "Afternoon practical session"
output: html_notebook
---


# GAMLSS modelling of aggregate UK electricity demand

- get the data and run a baseline GAM model


```{r}
library(mgcViz); library(gridExtra)
data("UKload")

form <- NetDemand~s(wM,k=20,bs='cr') + s(wM_s95,k=20,bs='cr') + 
        s(Posan, k=30, bs = "cc") + Dow + s(Trend,k=4,bs='cr') + 
        NetDemand.48 + Holy

fit0 <- gamV(form, data = UKload, aViz = list(nsim = 50)) 
```

- Look for patterns in the residuals conditional mean

```{r}
pl <- list()
pl[[1]] <- ( check1D(fit0, "wM") + l_gridCheck1D(gridFun = mean, stand = "sc") )$ggObj      # OK
pl[[2]] <- ( check1D(fit0, "wM_s95") + l_gridCheck1D(gridFun = mean, stand = "sc") )$ggObj  # OK
pl[[3]] <- ( check1D(fit0, "Posan") + l_gridCheck1D(gridFun = mean, stand = "sc") )$ggObj   # Not OK

grid.arrange(grobs = pl, ncol = 2)
```

- `Posan` has large absolute residuals in January and December, indicating that the cyclic smoothing may not be correct.

- Drop cyclic smoothing and recheck.

```{r}
form <- NetDemand ~ s(wM,k=20,bs='cr') + 
                    s(wM_s95,k=20,bs='cr') + 
                    s(Posan,bs='cr',k=30) +  # <- Changed `cc` to `cr`
                    Dow + s(Trend,k=4) + NetDemand.48 + Holy

fit1 <- gamV(form, data = UKload, aViz = list(nsim = 50)) 

# Pattern in residuals mean is gone!
check1D(fit1, "Posan") + l_gridCheck1D(gridFun = mean)
```

- Check the AIC

```{r}
AIC(fit0, fit1)
```

- Look at variable plots

```{r}
print(plot(fit1), pages = 1)
```

- Check the basis dimension basis

```{r}
tmp <- check(fit1)
```

- The `check` test indicates that k may be too low. Plotting the effect indicates that it drops sharply in December. 

- Fit a model with adaptive smoothing.

```{r}
form <- NetDemand ~ s(wM,k=20,bs='cr') + 
                    s(wM_s95,k=20,bs='cr') + 
                    s(Posan,bs='ad',k=30) +  # <- Changed `cr` to `ad`
                    Dow + s(Trend,k=4) + NetDemand.48 + Holy
fit2 <- gamV(form, data = UKload, aViz = list(nsim = 50)) 

AIC( fit1, fit2 )
```

- Plot the effect of `posan`

```{r}
plot(sm(fit2, 3), n = 400) + l_points() + l_fitLine() + l_ciLine() + theme_minimal()
```

- Now the mean model is finalised we can look at the conditional variance of residuals. Using density plots is helpful as it avoids just looking at the mean distribution.

```{r}
pl <- list()
pl[[1]] <- ( check1D(fit2, "wM") + l_densCheck(n = c(100, 100), tol = -1) )$ggObj
pl[[2]] <- ( check1D(fit2, "wM_s95") + l_densCheck(n = c(100, 100), tol = -1) )$ggObj
pl[[3]] <- ( check1D(fit2, "Posan") + l_densCheck(n = c(100, 100), tol = -1) )$ggObj

grid.arrange(grobs = pl, ncol = 2) 
```

- Some evidence of overdispersion check again using standardised plots

```{r}
pl <- list()
pl[[1]] <- ( check1D(fit2, "wM") + l_gridCheck1D(gridFun = sd, stand = "sc") )$ggObj
pl[[2]] <- ( check1D(fit2, "wM_s95") + l_gridCheck1D(gridFun = sd, stand = "sc") )$ggObj
pl[[3]] <- ( check1D(fit2, "Posan") + l_gridCheck1D(gridFun = sd, stand = "sc") )$ggObj

grid.arrange(grobs = pl, ncol = 2) # More evidence of heteroscedasticity
```

- All three variables having variable varinace. This could be addressed by fitting a GAMLSS model with variable scale. 

```{r}
form <- list(NetDemand ~ s(wM,k=20,bs='cr') + 
             s(wM_s95,k=20,bs='cr') + 
             s(Posan,bs='ad',k=30) + 
             Dow + s(Trend,k=4) + NetDemand.48 + Holy, 
             ~ s(wM_s95,k=10,bs='cr') + 
               s(Posan,bs='cr',k=20) + 
               Dow)

fit3 <- gamV(form, family = gaulss, data = UKload, aViz = list(nsim = 50))

AIC(fit2, fit3) 
```

- AIC has improved now check the variance.

```{r}
pl <- list()
pl[[1]] <- ( check1D(fit3, "wM") + l_gridCheck1D(gridFun = sd, stand = "sc") )$ggObj
pl[[2]] <- ( check1D(fit3, "wM_s95") + l_gridCheck1D(gridFun = sd, stand = "sc") )$ggObj
pl[[3]] <- ( check1D(fit3, "Posan") + l_gridCheck1D(gridFun = sd, stand = "sc") )$ggObj

grid.arrange(grobs = pl, ncol = 2) 
```

- Variance is reduced, relative to the location only Gaussian model. Now check skewness.

```{r}
qq(fit3)
```

- Fat tails and skewness to the left. Explore how this changes across covariates.

```{r}
library(e1071)
pl <- list()
pl[[1]] <- ( check1D(fit3, "wM_s95") + l_gridCheck1D(gridFun = skewness, stand = "sc") )$ggObj
pl[[2]] <- ( check1D(fit3, "Posan") + l_gridCheck1D(gridFun = skewness, stand = "sc") )$ggObj

grid.arrange(grobs = pl, ncol = 2)
```

- Residuals show departures from Guassian model based estimates (i.e response is symmetric). Fit a shash model to allow for skewed response

```{r}
library(mgcFam)
form <- list(NetDemand ~ s(wM,k=20,bs='cr') + 
               s(wM_s95,k=20,bs='cr') + 
               s(Posan,bs='ad',k=30) + 
               s(Trend,k=4) + NetDemand.48 + Holy + Dow, 
             ~ s(wM_s95,k=10,bs='cr') + 
               s(Posan,bs='cr',k=20) + 
               Dow, 
             ~ s(Posan, k = 10, bs='cr') + Dow, 
             ~ 1) # If convergence problems arise use
                  # ~ -1 + s(Holy, bs = "re", sp = 1e6) in place of ~ 1

fit4 <- gamV(form, family = shash, data = UKload, 
             aViz = list(nsim = 50))

AIC(fit3, fit4) # Decreased again by a lot
```

There is a clear improvement. Now recheck skewness.

```{r}
pl <- list()
pl[[1]] <- ( check1D(fit4, "wM_s95") + l_gridCheck1D(gridFun = skewness, stand = "sc") )$ggObj
pl[[2]] <- ( check1D(fit4, "Posan") + l_gridCheck1D(gridFun = skewness, stand = "sc") )$ggObj

grid.arrange(grobs = pl, ncol = 2) 
```

```{r}
print(plot(fit4, allTerms = TRUE), pages = 2, ask = F)
```

# Body Mass Index (BMI) of Dutch Boys

